# Forecasting Using R {-}

# Tutorial 1: Introduction to R {-}

```{r echo=FALSE, include=FALSE, message=FALSE}
library(ggplot2)
library(data.table)
```

R is a programming language for data analysis and visualisation. Here I introduce basic commands that should facilitate your understanding of R. You can further enhance your skillset using numerous online resources, as well as your own trial--and--error (which has been my own time-tested routine for mastering R). To the extent that new packages and features are added to R on a daily basis, there are virtually no limits to how far you can advance your knowledge of this programming language. 

We will work in RStudio---the go-to interface for R (as R itself is not an overly user-friendly platform). Thus, you will need to have installed both, R and RStudio on your devise (the latter will 'find' and connect with the former seamlessly). R is available from [CRAN](https://cran.r-project.org/){target="_blank"}, and RStudio is available from [RStudio](https://www.rstudio.com/){target="_blank"}. 

There are a number of ways in which we can work with data in R. To develop an intuition, let's start off with matrices. 

Consider a sequence of observations:
```{r echo=TRUE}
a <- c(1,0,4,3,2,6)
a
```

This sequence, unlike a *vector*, has no dimensions. But we can transform it to a $n \times 1$ vector using the `as.matrix()` function: 
```{r echo=TRUE}
b <- as.matrix(a)
b
```

The result is a $6 \times 1$ vector, or a column matrix. To obtain a $1 \times 6$ vector, or a row matrix, we *transpose* the aforementioned vector using the `t()` function:
```{r echo=TRUE}
bt <- t(b)
bt
dim(bt)
```

We can create any $n \times k$ matrix, using the `matrix()` function. For example, consider a $3 \times 2$ matrix:
```{r echo=TRUE}
B <- matrix(a,nrow=3,ncol=2)
B
```

We can add column names and row names to this matrix:
```{r echo=TRUE}
colnames(B) <- c("c1","c2")
rownames(B) <- c("r1","r2","r3")
B
```

If, at this point, we would like to only work with, say, the first column of the matrix, we can call it using its column number or the column name as follows:
```{r echo=TRUE}
B[,"c1"]
```

Similarly, if we want to refer to a matrix element, say $b_{3,2}$, we can do this by entering the respective indices in the brackets:
```{r echo=TRUE}
B[3,2]
```

Matrix multiplication is done using `%*%` command, granted that the two matrices are compatible. For example, we obtain a product of matrix $B$ and a new $2 \times 1$ vector, $d$, as follows:
```{r echo=TRUE}
d <- as.matrix(c(5,-2))
Bd <- B%*%d
Bd
```

We can add columns (and rows) to the existing matrix using a `cbind()` function:
```{r echo=TRUE}
c3 <- c(0,1,0)
D <- cbind(B,c3)
D
```

We can invert a(n invertible) matrix using the `solve()` function:
```{r echo=TRUE}
Di <- solve(D)
Di
```

One of the comparative advantages of R is in its graphing aesthetics. Currently, the best graphs are plotted via the **ggplot2** package. Notably, this package requires that the data are maintained in the `data.frame` or the `data.table` format (for the latter, you need to load the **data.table** package). Let's create a data.table object and observe its few lines:

```{r echo=TRUE}
set.seed(1)
x <- runif(120,0,2)
y <- 0.2+0.7*x+rnorm(120)

library(data.table)

dt <- data.table(y=y,x=x)
dt
```

Now, let's load **ggplot2** and generate a simple scatter plot:
```{r echo=TRUE, message=FALSE}
library(ggplot2)

ggplot(dt,aes(x=x,y=y))+
  geom_point()
```

We can augment this plot in a number of different ways. Here we change the point color to coral, add the fitted regression line to the plot, add labels to the figure, and apply a 'classic' background theme:
```{r echo=TRUE}
ggplot(dt,aes(x=x,y=y))+
  geom_point(color="coral")+
  geom_smooth(method="lm",formula=y~x,se=F,color="darkgray")+
  labs(title="A scatterplot with a fitted regression line", 
       x="Treatment Variable", 
       y="Outcome Variable", 
       caption="Caption: in case if needed.")+
  theme_classic()
```

As another example, let's generate a histogram (of the dependent variable):
```{r echo=TRUE}
ggplot(dt,aes(x=y))+
  geom_histogram(color="white",fill="coral",binwidth=.5)+
  labs(title="A basic histogram")+
  theme_classic()
```

We typically apply a line plot to illustrate a time series (that are ordered by date). In what follows, we add a date column to our data frame and then plot the dependent variable in the chronological order:
```{r echo=TRUE}
dt$date <- seq(from=as.Date("2000-01-01"),by="month",along.with=y)

ggplot(dt,aes(x=date,y=y))+
  geom_line(color="coral")+
  labs(title="A basic time series plot",  
       x="Year", 
       y="Outcome Variable")+
  theme_classic()
```

To illustrate the OLS regression in R, we apply the previously generated $x$ and $y$ as independent and dependent variables. To begin, we obtain the least squares estimator "by hand" as follows:
```{r echo=TRUE}
X <- cbind(1,x)
b <- solve(t(X)%*%X)%*%t(X)%*%y
b
```

This can be easily done using the `lm()` function:
```{r echo=TRUE}
ols <- lm(y~x)
ols
```

We can apply the `summary()` function to see the complete set of regression results:
```{r echo=TRUE}
summary(ols)
```


