# Tutorial 3: Forecasting Methods and Routines {-}

```{r echo=FALSE, include=FALSE, message=FALSE}
library(ggplot2)
library(data.table)
```

In this tutorial, we will introduce 'for loop', and use it to generate time series as well as to obtain one-step-ahead forecasts using a rolling window procedure; we will also perform forecast error diagnostics.

Let's generate a random walk process, such that $y_{t} = y_{t-1}+e_{t}$, where $e_{t} ~ N(0,1)$, and where $y_{0}=0$, for $t=1,\ldots,120$.

```{r echo=TRUE}
n <- 120

set.seed(1)
e <- rnorm(n)

y <- rep(NA,n)

y[1] <- e[1]

for(i in 2:n){
  y[i] <- y[i-1] + e[i]
}
```

Store $y$ and $e$ in a **data.table**, call it 'dt'. Add some arbitrary dates to the data (e.g., suppose we deal with the monthly series beginning from January 2011).

```{r echo=TRUE}
dt <- data.table(y,e)

dt$date <- seq(as.Date("2011-01-01"),as.Date("2020-12-01"),by="month")
```

Plot the realized time series using **ggplot** function.

```{r echo=TRUE}
ggplot(dt,aes(x=date,y=y))+
  geom_line(size=1)+
  labs(x="Year",y="Random Walk")+
  theme_classic()
```

Generate a sequence of one-step-ahead forecasts from naive and average methods, using the rolling window scheme, where the first rolling window ranges from period 1 to period 80.

```{r echo=TRUE}
dt$average <- NA
dt$naive <- NA

R <- 80
P <- n-R
for(i in 1:P){
  w <- y[i:(R-1+i)]
  dt$average[R+i] <- mean(w)
  dt$naive[R+i] <- w[length(w)]
}
```

Calculate the RMSFE measures for each of the two forecasting methods.

```{r echo=TRUE}
dt[,`:=`(e_average=y-average,e_naive=y-naive)]

rmsfe_average <- sqrt(mean(dt$e_average^2,na.rm=T))
rmsfe_naive <- sqrt(mean(dt$e_naive^2,na.rm=T))

rmsfe_average
rmsfe_naive
```

Perform the forecast error diagnostics for the two considered methods.

Zero mean of the forecast errors: $E(e_{t+1|t})=0$.
We perform this test by regressing the forecast error on the constant, and checking whether the coefficient is statistically significantly different from zero.
```{r echo=TRUE}
summary(lm(e_average~1,data=dt))$coefficients
summary(lm(e_naive~1,data=dt))$coefficients
```

No correlation of the forecast errors with the forecasts: $Cov(e_{t+1|t},y_{t+1|t})=0$. We perform this test by regressing the forecast error on the forecast, and checking whether the slope coefficient is statistically significantly different from zero.
```{r echo=TRUE}
summary(lm(e_average~average,data=dt))$coefficients
summary(lm(e_naive~naive,data=dt))$coefficients
```

No serial correlation in one-step-ahead forecast errors: $Cov(e_{t+1|t},y_{t|t-1})=0$. We perform this test by regressing the forecast error on its lag, and checking whether the slope coefficient is statistically significantly different from zero.
(Note: first we need to generate lagged forecast errors)
```{r echo=TRUE}
dt[,`:=`(e_average.l1=shift(e_average),e_naive.l1=shift(e_naive))]

summary(lm(e_average~e_average.l1,data=dt))$coefficients
summary(lm(e_naive~e_naive.l1,data=dt))$coefficients
```
