# Tutorial 4: Deterministic Trends {-}

```{r echo=FALSE, include=FALSE, message=FALSE}
library(ggplot2)
library(data.table)
```

In this tutorial, we will generate trending series, we will apply an information criterion to select the most suitable trend model, and we will obtain and compare one-step-ahead forecasts using a rolling window procedure.

Let's generate a time series that follows a quadratic trend: $y_{t} = 10+0.01t+0.002t^2+e_{t}$, where $e_{t} \sim N(0,16)$, for $t=1,\ldots,180$.

```{r echo=TRUE}
n <- 180

set.seed(7)
e <- rnorm(n,0,4)

trend <- c(1:n)

y <- 10+0.01*trend+0.002*trend^2+e
```

Store $y$ and $trend$ in a **data.table**, call it 'dt'. Add some arbitrary dates to the data (e.g., suppose we deal with the monthly series beginning from January 2006).

```{r echo=TRUE}
dt <- data.table(y,trend)

dt$date <- seq(as.Date("2006-01-01"),by="month",along.with=y)
```

Plot the realized time series using **ggplot** function.

```{r echo=TRUE}
ggplot(dt,aes(x=date,y=y))+
  geom_line(size=1)+
  labs(x="Year",y="Trending Series")+
  theme_classic()
```

Calculate Akaike Information Criteria for linear, quadratic, cubic, and exponential trend models, using all observations in the series.

```{r echo=TRUE}
AIC_vec <- matrix(ncol=4,nrow=1)
for(i in 1:4){
  if(i < 4){
    reg <- lm(y~poly(trend,degree=i,raw=T),data=dt)
    AIC_vec[i] <- log(crossprod(reg$residuals))+2*length(reg$coefficients)/n
  }else{
    reg <- lm(log(y)~trend,data=dt)
    yhat <- reg$fitted.values
    sig <- sd(reg$residuals)
    ystar <- exp(yhat+sig^2/2)
    res <- dt$y-ystar
    AIC_vec[i] <- log(crossprod(res))+2*length(reg$coefficients)/n
  }
}

AIC_vec
```

Generate a sequence of one-step-ahead forecasts from linear, quadratic, cubic, and exponential trend models, using the rolling window scheme, where the first rolling window ranges from period 1 to period 120.

```{r echo=TRUE}
dt$t1 <- NA
dt$t2 <- NA
dt$t3 <- NA
dt$te <- NA

R <- 120
P <- n-R
for(i in 1:P){
  reg1 <- lm(y~trend,data=dt[i:(R-1+i)])
  dt$t1[R+i] <- reg1$coef[1]+reg1$coef[2]*(R+i)
  
  reg2 <- lm(y~poly(trend,degree=2,raw=T),data=dt[i:(R-1+i)])
  dt$t2[R+i] <- reg2$coef[1]+reg2$coef[2]*(R+i)+reg2$coef[3]*((R+i)^2)
  
  reg3 <- lm(y~poly(trend,degree=3,raw=T),data=dt[i:(R-1+i)])
  dt$t3[R+i] <- reg3$coef[1]+reg3$coef[2]*(R+i)+reg3$coef[3]*((R+i)^2)+reg3$coef[4]*((R+i)^3)
  
  rege <- lm(log(y)~trend,data=dt[i:(R-1+i)])
  sig <- sd(rege$residuals)
  dt$te[R+i] <- exp(rege$coef[1]+rege$coef[2]*(R+i)+sig^2/2)
}
```

Plot the original series overlay by the one-step-ahead forecasts from the four considered trend models. Note, for convenience we will first 'melt' the data.table in to the 'long' format.

```{r echo=TRUE}
dt_long <- melt(dt[,.(date,y,linear=t1,quadratic=t2,cubic=t3,exponential=te)],id.vars="date")

ggplot(dt_long,aes(x=date,y=value,color=variable))+
  geom_line(size=1,na.rm=T)+
  scale_color_manual(values=c("darkgray","black","goldenrod","steelblue","indianred"))+
  labs(x="Year",y="Trending Series and Forecasts")+
  theme_classic()+
  theme(legend.title=element_blank(),legend.position=c(.15,.85))
```
  
Calculate the RMSFE measures for each of the two forecasting methods.

```{r echo=TRUE}
dt[,`:=`(e_t1=y-t1,e_t2=y-t2,e_t3=y-t3,e_te=y-te)]

rmsfe_t1 <- sqrt(mean(dt$e_t1^2,na.rm=T))
rmsfe_t2 <- sqrt(mean(dt$e_t2^2,na.rm=T))
rmsfe_t3 <- sqrt(mean(dt$e_t3^2,na.rm=T))
rmsfe_te <- sqrt(mean(dt$e_te^2,na.rm=T))

rmsfe_t1
rmsfe_t2
rmsfe_t3
rmsfe_te
```
